---
title: "analyze ODs"
author: "Catherine DiGennaro"
date: "5/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidycensus)
`%!in%` <- Negate(`%in%`)
#census_api_key("5512ff2f4dc8ad3053285afb019b439ed4bfab34", install=TRUE)
census_api_key(Sys.getenv("CENSUS_API_KEY"))
```

import OD data
```{r}
all_ODs <- read.csv("NCHS_all_ODs.csv") 
```

concatenate ICD columns
```{r}
all_ODs <- all_ODs %>% 
  unite(all_codes, record_1:record_20, sep = "|", remove=TRUE)
```

assign specific drug category codes 
```{r}
# opioids <- "T400|T401|T402|T403|T404|T406"
# opioids_without_fentanyl <- "T400|T401|T402|T403|T406"
heroin <- "T401"
rx_opioids <- "T400|T402|T403"
other <- "T406"
fentanyl <- "T404"
cocaine <- "T405"
psychostimulants <- "T436"
# alcohol <- "E244|F10|G312|G621|G721|I426|K292|K70|K852|K860|R780|X45|X65|Y15"
```

classify each overdose by type of drug present
```{r}
ODs_classified <- all_ODs %>%
  mutate(
    heroin = case_when(grepl(heroin, all_codes) ~ 1,
                       TRUE ~ 0),
    rx_opioids = case_when(grepl(rx_opioids, all_codes) ~ 1,
                           TRUE ~ 0),
    fentanyl = case_when(grepl(fentanyl, all_codes) ~ 1,
                         TRUE ~ 0),
    other = case_when(grepl(other, all_codes) ~ 1,
                      TRUE ~ 0),
    cocaine = case_when(grepl(cocaine, all_codes) ~ 1,
                        TRUE ~ 0),
    psychostimulants = case_when(grepl(psychostimulants, all_codes) ~ 1,
                                 TRUE ~ 0)
    # ,alcohol = case_when(grepl(alcohol, all_codes) ~ 1,
    #                      TRUE ~ 0))
  )

```

skip for FDA grant;
classify each drug overdose by whether alcohol was present
```{r}
ODs_alcohol <- ODs_classified %>% 
  mutate(opioids_w_alc = case_when(opioids==1 &
                                     alcohol==1 ~ 1,
                                   TRUE ~ 0),
         opioids_wo_alc = case_when(opioids==1 &
                                     alcohol==0 ~ 1,
                                   TRUE ~ 0),
         opioids_without_fentanyl_w_alc = case_when(opioids_without_fentanyl==1 &
                                     alcohol==1 ~ 1,
                                   TRUE ~ 0),
         opioids_without_fentanyl_wo_alc = case_when(opioids_without_fentanyl==1 &
                                     alcohol==0 ~ 1,
                                   TRUE ~ 0),
         fentanyl_w_alc = case_when(fentanyl==1 &
                                     alcohol==1 ~ 1,
                                   TRUE ~ 0),
         fentanyl_wo_alc = case_when(fentanyl==1 &
                                     alcohol==0 ~ 1,
                                   TRUE ~ 0),
         cocaine_w_alc = case_when(cocaine==1 &
                                     alcohol==1 ~ 1,
                                   TRUE ~ 0),
         cocaine_wo_alc = case_when(cocaine==1 &
                                     alcohol==0 ~ 1,
                                   TRUE ~ 0),
         psychostimulants_w_alc = case_when(psychostimulants==1 &
                                     alcohol==1 ~ 1,
                                   TRUE ~ 0),
         psychostimulants_wo_alc = case_when(psychostimulants==1 &
                                     alcohol==0 ~ 1,
                                   TRUE ~ 0))
```


assign mutually exclusive labels to drug overdoses
```{r}
ODs_mut_excl <- ODs_classified %>%
  mutate(
    heroin_only = case_when(
      heroin == 1 &
        rx_opioids == 0 &
        fentanyl == 0 &
        cocaine == 0 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
    ),
    rx_opioids_only = case_when(
      heroin == 0 &
        rx_opioids == 1 &
        fentanyl == 0 &
        cocaine == 0 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
    ),
    fentanyl_only = case_when(
      heroin == 0 &
        rx_opioids == 0 &
        fentanyl == 1 &
        cocaine == 0 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
    ),
    cocaine_only = case_when(
      heroin == 0 &
        rx_opioids == 0 &
        fentanyl == 0 &
        cocaine == 1 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
    ),
    psychostimulants_only = case_when(
      heroin == 0 &
        rx_opioids == 0 &
        fentanyl == 0 &
        cocaine == 0 &
        psychostimulants == 1 ~ 1,
      TRUE ~ 0
    ))
```

assign polysubstance drug overdoses by combos of drugs present
```{r}
ODs_polysub <- ODs_mut_excl %>%
  mutate(
    heroin_with_or_without_rx = case_when(
      heroin == 1 &
        fentanyl == 0 &
        cocaine == 0 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
    ),
    PS_and_synthetics_with_or_without_opioids_no_coc = case_when(
      psychostimulants == 1 &
        fentanyl == 1 &
        cocaine == 0 ~ 1,
      TRUE ~ 0
    ),
    coc_and_synthetics_with_or_without_opioids_no_PS = case_when(
      psychostimulants == 0 &
        fentanyl == 1 &
        cocaine == 1 ~ 1,
      TRUE ~ 0
    ),
    coc_with_heroin_or_rx = case_when(
      cocaine == 1 &
        (heroin == 1 |
        rx_opioids == 1) &
        fentanyl == 0 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
    ),
    ps_with_heroin_or_rx = case_when(
      cocaine == 0 &
        (heroin == 1 |
        rx_opioids == 1) &
        fentanyl == 0 &
        psychostimulants == 1 ~ 1,
      TRUE ~ 0
    ),
    synthetics_without_PS_or_coc_with_or_without_other_opioids = case_when(
      fentanyl == 1 &
        psychostimulants == 0 &
        cocaine == 0 ~ 1,
      TRUE ~ 0
    ),
    PS_coc_with_or_without_opioids = case_when(
      psychostimulants == 1 &
        cocaine == 1 ~ 1,
      TRUE ~ 0
    ),
    # PS_coc_synthetics_with_or_without_rx_and_heroin = case_when(
    #   psychostimulants == 1 &
    #     cocaine == 1 &
    #     fentanyl == 1  ~ 1,
    #   TRUE ~ 0
    # ),
    # PS_coc_only = case_when(
    #   psychostimulants == 1 &
    #     cocaine == 1 &
    #     fentanyl == 0 &
    #     rx_opioids == 0 &
    #     heroin == 0 ~ 1,
    #   TRUE ~ 0
    # ),
    all_other_combos = case_when(
      rx_opioids_only == 0 &
        heroin_with_or_without_rx == 0 &
        cocaine_only == 0 &
        psychostimulants_only == 0 &
        PS_and_synthetics_with_or_without_opioids_no_coc == 0 &
        coc_and_synthetics_with_or_without_opioids_no_PS == 0 &
        coc_with_heroin_or_rx == 0 &
        ps_with_heroin_or_rx == 0 &
        synthetics_without_PS_or_coc_with_or_without_other_opioids == 0 &
        PS_coc_with_or_without_opioids == 0 ~ 1,
      TRUE ~ 0
    ),
    ps_or_coc_NOT_MUTEXCL = case_when(
      (psychostimulants == 1 |
        cocaine == 1 ~ 1),
      TRUE ~ 0
    )
    
    #,
    # ps_and_cocaine_only = case_when(cocaine == 1 &
    #                                   opioids == 0 &
    #                                   psychostimulants == 1 ~ 1,
    #                                 TRUE ~ 0),
    # opioids_ps_and_cocaine = case_when(cocaine == 1 &
    #                                      opioids == 1 &
    #                                      psychostimulants == 1 ~ 1,
    #                                    TRUE ~ 0),
    # opioids_without_fentanyl_and_cocaine_only = case_when(
    #   cocaine == 1 &
    #     opioids_without_fentanyl == 1 &
    #     fentanyl == 0 &
    #     psychostimulants == 0 ~ 1,
    #   TRUE ~ 0
    # ),
    # opioids_without_fentanyl_and_ps_only = case_when(
    #   cocaine == 0 &
    #     opioids_without_fentanyl == 1 &
    #     fentanyl == 0 &
    #     psychostimulants == 1 ~ 1,
    #   TRUE ~ 0
    # ),
    # opioids_without_fentanyl_ps_and_cocaine = case_when(
    #   cocaine == 1 &
    #     opioids_without_fentanyl == 1 &
    #     fentanyl == 0 &
    #     psychostimulants == 1 ~ 1,
    #   TRUE ~ 0
    # ),
    # fentanyl_and_cocaine_only = case_when(
    #   cocaine == 1 &
    #     fentanyl == 1 &
    #     opioids_without_fentanyl ==
    #     0 &
    #     psychostimulants == 0 ~ 1,
    #   TRUE ~ 0
    # ),
    # fentanyl_and_ps_only = case_when(
    #   cocaine == 0 &
    #     fentanyl == 1 &
    #     opioids_without_fentanyl == 0 &
    #     psychostimulants == 1 ~ 1,
    #   TRUE ~ 0
    # ),
    # fentanyl_ps_and_cocaine = case_when(
    #   cocaine == 1 &
    #     fentanyl == 1 &
    #     opioids_without_fentanyl == 0 &
    #     psychostimulants == 1 ~ 1,
    #   TRUE ~ 0
    # )
  )
```


generate summaries by year and save to output file
```{r}

# OD_summary <- ODs_polysub %>% 
#   group_by(year) %>% 
#   summarize("Rx Opioids Only" = sum(rx_opioids_only),
#             "Fentanyl Only" = sum(fentanyl_only),
#             "Cocaine Only"= sum(cocaine_only),
#             "Psychostimulants Only" = sum(psychostimulants_only),
#             "Heroin with or without Rx" = sum(heroin_with_or_without_rx),
            
            
  
  
OD_summary_not_mu <- ODs_polysub %>% 
  group_by(year) %>% 
  summarize("Opioid Overdose Deaths" = sum(opioids),
            "Opioid Overdose Deaths without Fentanyl" = sum(opioids_without_fentanyl),
            "Fentanyl Overdose Deaths" = sum(fentanyl),
            "Cocaine Overdose Deaths" = sum(cocaine),
            "Psychostimulant Overdose Deaths" = sum(psychostimulants))

OD_summary_mu <- ODs_polysub %>% 
  group_by(year) %>% 
  summarize("Opioid-only Overdose Deaths" = sum(opioids_only),
            "Opioid-only Overdose Deaths without Fentanyl" = sum(opioids_without_fentanyl_only),
            "Fentanyl-only Overdose Deaths" = sum(fentanyl_only),
            "Cocaine-only Overdose Deaths" = sum(cocaine_only),
            "Psychostimulant-only Overdose Deaths" = sum(psychostimulants_only))

OD_summary_not_mu_alc <- ODs_polysub %>% 
  group_by(year) %>% 
  summarize("Opioid Overdose Deaths with Alcohol" = sum(opioids_w_alc),
            "Opioid Overdose Deaths without Alcohol" = sum(opioids_wo_alc),
            "Opioid Overdose Deaths without Fentanyl with Alcohol" = sum(opioids_without_fentanyl_w_alc),
            "Opioid Overdose Deaths without Fentanyl without Alcohol" = sum(opioids_without_fentanyl_wo_alc),
            "Fentanyl Overdose Deaths with Alcohol" = sum(fentanyl_w_alc),
            "Fentanyl Overdose Deaths without Alcohol" = sum(fentanyl_wo_alc),
            "Cocaine Overdose Deaths with Alcohol" = sum(cocaine_w_alc),
            "Cocaine Overdose Deaths without Alcohol" = sum(cocaine_wo_alc),
            "Psychostimulant Overdose Deaths with Alcohol" = sum(psychostimulants_w_alc),
            "Psychostimulant Overdose Deaths without Alcohol" = sum(psychostimulants_wo_alc))

OD_summary_mu_alc <- ODs_polysub %>% 
  group_by(year) %>% 
  summarize("Opioid-only Overdose Deaths with Alcohol" = sum(opioids_only_w_alc),
            "Opioid-only Overdose Deaths without Alcohol" = sum(opioids_only_wo_alc),
            "Opioid-only Overdose Deaths without Fentanyl with Alcohol" = sum(opioids_without_fentanyl_only_w_alc),
            "Opioid-only Overdose Deaths without Fentanyl without Alcohol" = sum(opioids_without_fentanyl_only_wo_alc),
            "Fentanyl-only Overdose Deaths with Alcohol" = sum(fentanyl_only_w_alc),
            "Fentanyl-only Overdose Deaths without Alcohol" = sum(fentanyl_only_wo_alc),
            "Cocaine-only Overdose Deaths with Alcohol" = sum(cocaine_only_w_alc),
            "Cocaine-only Overdose Deaths without Alcohol" = sum(cocaine_only_wo_alc),
            "Psychostimulant-only Overdose Deaths with Alcohol" = sum(psychostimulants_only_w_alc),
            "Psychostimulant-only Overdose Deaths without Alcohol" = sum(psychostimulants_only_wo_alc))

OD_summary_polysub <- ODs_polysub %>% 
  group_by(year) %>% 
  summarize("Opioid and Cocaine Overdose without Psychostimulants" = sum(opioids_and_cocaine_only),
            "Opioid and Psychostimulant Overdose without Cocaine" = sum(opioids_and_ps_only),
            "Psychostimulant and Cocaine Overdose without Opioids" = sum(ps_and_cocaine_only),
            "Opioid Overdose with Cocaine OR Psychostimulants" = sum(opioids_and_ps_or_cocaine),
            "Opioid, Psychostimulant, and Cocaine Overdoses" = sum(opioids_ps_and_cocaine),
            "Opioid and Cocaine Overdose without Fentanyl or Psychostimulants" = sum(opioids_without_fentanyl_and_cocaine_only),
            "Opioid and Psychostimulant Overdose without Fentanyl or Cocaine" = sum(opioids_without_fentanyl_and_ps_only),
            "Opioid, Psychostimulant, and Cocaine Overdose without Fentanyl" = sum(opioids_without_fentanyl_ps_and_cocaine),
            "Fentanyl and Cocaine Overdose without other opioids or Psychostimulants" = sum(fentanyl_and_cocaine_only),
            "Fentanyl and Psychositimulant Overdose without other opioids or Cocaine" = sum(fentanyl_and_ps_only),
            "Fentanyl, Psychostimulant, and Cocaine Overdose without other opioids" = sum(fentanyl_ps_and_cocaine))

all_summaries <- cbind(OD_summary_not_mu, 
                       OD_summary_mu[,-1], 
                       OD_summary_not_mu_alc[,-1], 
                       OD_summary_mu_alc[,-1],
                       OD_summary_polysub[,-1])
#write.csv(all_summaries, "OD_category_summaries.csv")
```

summary needed for FDA grant
```{r}
grant_summary <- ODs_polysub %>%
  group_by(year) %>%
  summarize(
    "Rx opioids only" = sum(rx_opioids_only),
    "Heroin with or without Rx opioids" = sum(heroin_with_or_without_rx),
    "Cocaine only" = sum(cocaine_only),
    "Psychostimulants only" = sum(psychostimulants_only),
    "Psychostimulants and synthetics with or without other opioids" = sum(PS_and_synthetics_with_or_without_opioids_no_coc),
    "Cocaine and synthetics with or without other opioids" = sum(coc_and_synthetics_with_or_without_opioids_no_PS),
    "Cocaine with heroin and/or Rx opioids" = sum(coc_with_heroin_or_rx),
    "Psychostimulants with heroin and/or Rx opioids" = sum(ps_with_heroin_or_rx),
    "Synthetics with or without other opioids" = sum(synthetics_without_PS_or_coc_with_or_without_other_opioids),
    "Psychostimulants and cocaine with or without opioids" = sum(PS_coc_with_or_without_opioids),
    "All other drug combinations" = sum(all_other_combos)
  )

write.csv(grant_summary, "summary_ODs.csv")

```

categorize OD deaths for model
```{r}
ODs_model <- ODs_classified %>% 
  filter(heroin == 1 | rx_opioids == 1 | fentanyl == 1 | other == 1) %>% 
  mutate(
    rx_only = case_when(
      heroin == 0 &
        rx_opioids == 1 &
        other == 0 &
        fentanyl == 0 ~ 1,
      TRUE ~ 0
    ),
    rx_and_synth = case_when(
      heroin == 0 &
        rx_opioids == 1 &
        other == 0 &
        fentanyl == 1 ~ 1,
      TRUE ~ 0
    ),
    rx_and_heroin = case_when(
      heroin == 1 &
        rx_opioids == 1 &
        other == 0 &
        fentanyl == 0 ~ 1,
      TRUE ~ 0
    ),
    rx_and_other = case_when(
      heroin == 0 &
        rx_opioids == 1 &
        other == 1 &
        fentanyl == 0 ~ 1,
      TRUE ~ 0
    ),
    rx_multiple = case_when(  #category 5
      heroin == 0 &
        rx_opioids == 1 &
        other == 1 &
        fentanyl == 1 ~ 1, 
      TRUE ~ 0
    ),
    heroin_only = case_when(
      heroin == 1 &
        rx_opioids == 0 &
        other == 0 &
        fentanyl == 0 ~ 1,
      TRUE ~ 0
    ),
    heroin_with_synth = case_when(
      heroin == 1 &
        rx_opioids == 0 &
        other == 0 &
        fentanyl == 1 ~ 1,
      TRUE ~ 0
    ),
    heroin_with_other = case_when(
      heroin == 1 &
        rx_opioids == 0 &
        other == 1 &
        fentanyl == 0 ~ 1,
      TRUE ~ 0
    ),
    heroin_multiple = case_when(  #Category 9
      heroin == 1 &
        ((rx_opioids == 1 & other == 1) |
           (rx_opioids == 1 & fentanyl == 1) |
           (fentanyl == 1 & other == 1)) ~ 1, 
    TRUE ~ 0
    ),
    synth_only = case_when(
      heroin == 0 &
        rx_opioids == 0 &
        other == 0 &
        fentanyl == 1 ~ 1,
      TRUE ~ 0
    ),
    synth_other = case_when(
      heroin == 0 &
        rx_opioids == 0 &
        other == 1 &
        fentanyl == 1 ~ 1,
      TRUE ~ 0
    ),
    other_only = case_when(
      heroin == 0 &
        rx_opioids == 0 &
        other == 1 &
        fentanyl == 0 ~ 1,
      TRUE ~ 0
    ))
```


summary needed for model
```{r}
model_summary <- ODs_model %>% 
  group_by(year) %>% 
  summarize("(1) rx opioids only" = sum(rx_only),
            "(2) rx opioids w/ synth" = sum(rx_and_synth),
            "(3) rx opioids w/ heroin" = sum(rx_and_heroin),
            "(4) rx opioids w/ other" = sum(rx_and_other),
            "(5) rx multiple" = sum(rx_multiple),
            "(6) heroin only" = sum(heroin_only),
            "(7) heroin w/ synth" = sum(heroin_with_synth),
            "(8) heroin w/ other" = sum(heroin_with_other),
            "(9) heroin multiple" = sum(heroin_multiple),
            "(10) synth only" = sum(synth_only),
            "(11) synth w/ other" = sum(synth_other),
            "(12) other only" = sum(other_only)
    
  )

write.csv(model_summary, "summary_nvss_overdose_deaths.csv")
```


supplementary quick analyses ---

PS quick summary - how many psychostimulant-involved ODs are psychostimulant-only?
```{r}
ODs_psychostimulants <- ODs_polysub %>%
  mutate(any_PS = case_when(psychostimulants == 1 ~ 1,
                            TRUE ~ 0))

PS_summary <- ODs_psychostimulants %>% 
  group_by(year) %>% 
  summarize(
    "All psychostimulant-involved overdoses" = sum(any_PS),
    "Psychostimulant-only overdoses" = sum(psychostimulants_only),
    "Prop psychostimulant-only ODs out of all psychostimulant ODs" = sum(psychostimulants_only)/sum(any_PS)
  )

ps_p <- PS_summary %>% 
  ggplot(aes(x=year, y=`Prop psychostimulant-only ODs out of all psychostimulant ODs`)) + 
  geom_line(color='blue') + 
  coord_cartesian(ylim = c(0, .7)) +
  theme_bw()

ps_p
```

all possible combinations
```{r}
drugs <- c("rx_opioids", "heroin", "fentanyl", "psychostimulants", "cocaine")
two <- data.frame(combn(drugs, 2))
three <- data.frame(combn(drugs, 3))
four <- data.frame(combn(drugs, 4))

# write.csv(two, "two.csv")
# write.csv(three, "three.csv")
# write.csv(four, "four.csv")
```

summary of cocaine or PS, not mutually exclusive
```{r}
ps_or_coc <- ODs_polysub %>% 
  group_by(year) %>% 
  summarize("Psychostimulant- or Cocaine-involved Overdoses" = sum(ps_or_coc_NOT_MUTEXCL))

write.csv(ps_or_coc, "PS_or_Cocaine_ODs.csv")

ps_OR_coc_p <- ps_or_coc %>% 
  ggplot(aes(x = year, y = `Psychostimulant- or Cocaine-involved Overdoses`)) + 
  geom_line() + 
  theme_bw()
```

how many cocaine and meth involved deaths involved fent with other opioids, fent only, and other opioids but not fent?
```{r}
ODs_coc_and_meth <- ODs_classified %>% 
   mutate(
     coc_meth_fent_other_opioids = case_when(
      (heroin == 1 | rx_opioids == 1) &
        fentanyl == 1 &
        (cocaine == 1 | psychostimulants == 1) ~ 1,
      TRUE ~ 0
     ),
     coc_fent = case_when(
       heroin == 0 & 
        rx_opioids == 0 &
        fentanyl == 1 &
        cocaine == 1 &
        psychostimulants == 0 ~ 1,
      TRUE ~ 0
     ),
     meth_fent = case_when(
       heroin == 0 & 
        rx_opioids == 0 &
        fentanyl == 1 &
        cocaine == 0 &
        psychostimulants == 1 ~ 1,
      TRUE ~ 0
     ),
     coc_and_meth_and_fent = case_when(
       heroin == 0 & 
        rx_opioids == 0 &
        fentanyl == 1 &
        cocaine == 1 &
        psychostimulants == 1 ~ 1,
      TRUE ~ 0
     ),
     # coc_meth_fent_only = case_when(
     #  heroin == 0 & 
     #    rx_opioids == 0 &
     #    fentanyl == 1 &
     #    (cocaine == 1 | psychostimulants == 1) ~ 1,
     #  TRUE ~ 0
     # ),
     coc_meth_other_opioids = case_when(
      (heroin == 1 | rx_opioids == 1) &
        fentanyl == 0 &
        (cocaine == 1 | psychostimulants == 1) ~ 1,
      TRUE ~ 0
     ),
     coc_meth_no_opioids = case_when(
       heroin == 0 & 
        rx_opioids == 0 &
        fentanyl == 0 &
        (cocaine == 1 | psychostimulants == 1) ~ 1,
      TRUE ~ 0
     ),
   )
```

summarize cocaine and meth data
```{r}
coc_and_meth <- ODs_coc_and_meth %>% 
  group_by(year) %>% 
  summarize("Cocaine or Meth with Fent and Other Opioids" = sum(coc_meth_fent_other_opioids),
            "Cocaine with Fentanyl without Other Opioids"= sum(coc_fent),
            "Meth with Fentanyl without Other Opioids" = sum(meth_fent),
            "Cocaine and Meth with Fentanyl no Other Opioids" = sum(coc_and_meth_and_fent),
            #"Cocaine or Meth with Fent and not Other Opioids" = sum(coc_meth_fent_only),
            "Cocaine or Meth with Other Opioids and not Fent" = sum(coc_meth_other_opioids),
            "Cocaine or Meth without Opioids" = sum(coc_meth_no_opioids))

write.csv(coc_and_meth, "NVSS_cocaine_and_meth.csv")
  
```

Demographic analysis - race
```{r}
#census race variables
race_vars <- c(
Hispanic = "B03001_003",
`NH White` = "B03002_003",
`NH Black` = "B03002_004",
`NH Asian` = "B03002_006",
`NH Native American` = "B03002_005"
)

race_by_yr <- get_acs(geography = "us", variables = race_vars, year = 2020)

ODs_race <- ODs_polysub %>% 
  mutate(race_recode = case_when(
    (hispanic == 0 | between(hispanic, 100, 199) | between(hispanic, 996, 999) | hispanic == 6) & race == 1 ~ "NH White",
    (hispanic == 0 | between(hispanic, 100, 199) | between(hispanic, 996, 999) | hispanic == 7) & race == 2 ~ "NH Black", 
    (hispanic == 0 | between(hispanic, 100, 199) | between(hispanic, 996, 999)) & race == 3 ~ "NH Native American",
    (hispanic == 0 | between(hispanic, 100, 199) | between(hispanic, 996, 999)) & race %in% c(4, 5, 6, 7, 18, 28, 38, 48, 58, 68, 78) ~ "NH Asian",
    (hispanic == 1 | between(hispanic, 200, 299) | hispanic %in% c(1, 2, 3,4, 5, 9, 99))  ~ "Hispanic",
    TRUE ~ "ERROR"
  )) 

OD_race_summary <- ODs_race %>% 
  group_by(year, race_recode) %>% 
  summarize("Heroin-involved" = sum(heroin),
            "Rx opioids-involved" = sum(rx_opioids),
            "Fentanyl-involved" = sum(fentanyl),
            "Other opioids-involved" = sum(other),
            "Cocaine-involved" = sum(cocaine),
            "PS-involved" = sum(psychostimulants),
            "Rx opioids only" = sum(rx_opioids_only), 
            "Fentanyl only" = sum(fentanyl_only),
            "Cocaine only" =  sum(cocaine_only),
            "PS only" = sum(psychostimulants_only),
            "Heroin with or without Rx" = sum(heroin_with_or_without_rx),
            "PS and fentanyl, no cocaine" = sum(PS_and_synthetics_with_or_without_opioids_no_coc),
            "Cocaine and fentanyl, no PS" = sum(coc_and_synthetics_with_or_without_opioids_no_PS),
            "Cocaine with heroin or Rx" = sum(coc_with_heroin_or_rx),
            "PS with heroin or Rx" = sum(ps_with_heroin_or_rx),
            "Fentanyl without stimulants, with or without opioids" = sum(synthetics_without_PS_or_coc_with_or_without_other_opioids),
            "PS and Cocaine, with or without opioids" = sum(PS_coc_with_or_without_opioids),
            "All other combos" = sum(all_other_combos)
            ) %>% 
  pivot_longer(!c(year, race_recode), names_to = "drug", values_to = "num_deaths") %>% 
  left_join(race_by_yr, by = c("race_recode" = "variable")) %>% 
  mutate(std_prop_race = num_deaths/estimate)

race_p <- OD_race_summary %>% 
  filter(drug %!in% c("Heroin-involved", "Rx opioids-involved", "Fentanyl-involved", "Other opioids-involved", "Cocaine-involved", "PS-involved")) %>%
  ggplot(aes(x = year, y = std_prop_race, color = drug)) +
  geom_line() + 
  labs(y = "Overdose Death Rate", color = "Drug") +
  facet_wrap(~race_recode) +
  theme_bw() + 
  theme(legend.position = "bottom")

simple_drugs_race_p <- OD_race_summary %>% 
  filter(drug %in% c("Heroin-involved", "Rx opioids-involved", "Fentanyl-involved", "Other opioids-involved", "Cocaine-involved", "PS-involved")) %>%  
  ggplot(aes(x = year, y = std_prop_race, color = drug)) +
  geom_line() +
  labs(y = "Overdose Death Rate", color = "Drug Involved") +
  scale_y_continuous(labels = number) +
  facet_wrap(~race_recode) +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(plot = race_p, filename = "race_p.svg", device= "svg", height = 7, width = 11)
ggsave(plot = simple_drugs_race_p, filename = "simple_drugs_race_p.svg", device= "svg", height = 7, width = 11)
```

Demographic analysis - age
```{r}
ODs_age <- ODs_polysub %>% 
  mutate(age_recode = case_when(
    ager12 == 1 ~ "Under 1 year",
    ager12 == 2 ~ "1-4 years",
    ager12 == 3 ~ "5-14 years",
    ager12 == 4 ~ "15-24 years",
    ager12 == 5 ~ "25-34 years", 
    ager12 == 6 ~ "35-44 years",
    ager12 == 7 ~ "45-54 years",
    ager12 == 8 ~ "55-64 years",
    ager12 == 9 ~ "65-74 years",
    ager12 == 10 ~ "75-84 years",
    ager12 == 11 ~ "85+ years",
    ager12 == 12 ~ "Age unknown",
    TRUE ~ "ERROR"
  ))

OD_age_summary <- ODs_age %>% 
  group_by(year, age_recode) %>% 
  summarize("Heroin-involved" = sum(heroin),
            "Rx opioids-involved" = sum(rx_opioids),
            "Fentanyl-involved" = sum(fentanyl),
            "Other opioids-involved" = sum(other),
            "Cocaine-involved" = sum(cocaine),
            "PS-involved" = sum(psychostimulants),
            "Rx opioids only" = sum(rx_opioids_only), 
            "Fentanyl only" = sum(fentanyl_only),
            "Cocaine only" =  sum(cocaine_only),
            "PS only" = sum(psychostimulants_only),
            "Heroin with or without Rx" = sum(heroin_with_or_without_rx),
            "PS and fentanyl, no cocaine" = sum(PS_and_synthetics_with_or_without_opioids_no_coc),
            "Cocaine and fentanyl, no PS" = sum(coc_and_synthetics_with_or_without_opioids_no_PS),
            "Cocaine with heroin or Rx" = sum(coc_with_heroin_or_rx),
            "PS with heroin or Rx" = sum(ps_with_heroin_or_rx),
            "Fentanyl without stimulants, with or without opioids" = sum(synthetics_without_PS_or_coc_with_or_without_other_opioids),
            "PS and Cocaine, with or without opioids" = sum(PS_coc_with_or_without_opioids),
            "All other combos" = sum(all_other_combos)
            ) %>% 
  pivot_longer(!c(year, age_recode), names_to = "drug", values_to = "num_deaths")

age_p <- OD_age_summary %>% 
  filter(drug %!in% c("Heroin-involved", "Rx opioids-involved", "Fentanyl-involved", "Other opioids-involved", "Cocaine-involved", "PS-involved"),
         age_recode %in% c("15-24 years", "25-34 years", "35-44 years", "45-54 years", "55-64 years", "65-74 years", "75-84 years", "85+ years")) %>%
  ggplot(aes(x = year, y = num_deaths, color = drug)) +
  geom_line() + 
  facet_wrap(~age_recode) +
  theme_bw()

simple_drugs_age_p <- OD_age_summary %>% 
  filter(drug %in% c("Heroin-involved", "Rx opioids-involved", "Fentanyl-involved", "Other opioids-involved", "Cocaine-involved", "PS-involved"),
         age_recode %in% c("15-24 years", "25-34 years", "35-44 years", "45-54 years", "55-64 years", "65-74 years", "75-84 years", "85+ years")) %>%  
  ggplot(aes(x = year, y = num_deaths, color = drug)) +
  geom_line() + 
  facet_wrap(~age_recode) +
  theme_bw()
```
Demographic analysis - sex
```{r}
ODs_sex <- ODs_polysub %>% 
  mutate(sex_recode = case_when(
    sex == 1 | sex == "M" ~ "Male",
    sex == 2 | sex == "F" ~ "Female",
    TRUE ~ "ERROR"
  ))

OD_sex_summary <- ODs_sex %>% 
  group_by(year, sex_recode) %>% 
  summarize("Heroin-involved" = sum(heroin),
            "Rx opioids-involved" = sum(rx_opioids),
            "Fentanyl-involved" = sum(fentanyl),
            "Other opioids-involved" = sum(other),
            "Cocaine-involved" = sum(cocaine),
            "PS-involved" = sum(psychostimulants),
            "Rx opioids only" = sum(rx_opioids_only), 
            "Fentanyl only" = sum(fentanyl_only),
            "Cocaine only" =  sum(cocaine_only),
            "PS only" = sum(psychostimulants_only),
            "Heroin with or without Rx" = sum(heroin_with_or_without_rx),
            "PS and fentanyl, no cocaine" = sum(PS_and_synthetics_with_or_without_opioids_no_coc),
            "Cocaine and fentanyl, no PS" = sum(coc_and_synthetics_with_or_without_opioids_no_PS),
            "Cocaine with heroin or Rx" = sum(coc_with_heroin_or_rx),
            "PS with heroin or Rx" = sum(ps_with_heroin_or_rx),
            "Fentanyl without stimulants, with or without opioids" = sum(synthetics_without_PS_or_coc_with_or_without_other_opioids),
            "PS and Cocaine, with or without opioids" = sum(PS_coc_with_or_without_opioids),
            "All other combos" = sum(all_other_combos)
            ) %>% 
  pivot_longer(!c(year, sex_recode), names_to = "drug", values_to = "num_deaths")

sex_p <- OD_sex_summary %>% 
  filter(drug %!in% c("Heroin-involved", "Rx opioids-involved", "Fentanyl-involved", "Other opioids-involved", "Cocaine-involved", "PS-involved")) %>%
  ggplot(aes(x = year, y = num_deaths, color = drug)) +
  geom_line() + 
  facet_wrap(~sex_recode) +
  theme_bw()

simple_drugs_sex_p <- OD_sex_summary %>% 
  filter(drug %in% c("Heroin-involved", "Rx opioids-involved", "Fentanyl-involved", "Other opioids-involved", "Cocaine-involved", "PS-involved")) %>%  
  ggplot(aes(x = year, y = num_deaths, color = drug)) +
  geom_line() + 
  facet_wrap(~sex_recode) +
  theme_bw()
```
 

